Probabilistic Forecasting of Geomagnetic Indices using Gaussian Process Models

### 2017-07-21 revised-chapter.tex
Revisions begun

In this chapter, we give the reader an indepth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters. We highlight design decisions and practical issues must be addressed in order to use \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest. As a paedadogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.

@@

The Earth's magnetosphere is a region around the planet where the Earth's own magnetic field dominates the magnetic field of outer space. It is a region which is impinged upon constantly by the solar wind. The ionised plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere. It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.

@@

    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}. A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure. 
\end{enumerate}

Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forcasts for various geomagnetic time series. The principal data source for forecasting of geomagnetic indices is the \emph{OMNI} data set hosted by NASA on the \emph{Space Physics Data Facility} (SPDF).

@@

Probabilistic forecasting is of particular importance in geophysics applications as the end users of forecasts often require confidence bounds on the said forecasts. It is in this context where \emph{Gaussian Processes} become especially attractive due to their inherent probabilistic formulation and tractability of analytical inference.

@@

In the machine learning community, $m(.)$ and $K(.,.)$ are known as the \emph{mean function} and \emph{covariance function} or \emph{kernel function} of the process respectively. Giving a closed form expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed notationally as shown below.

>>>>

In this chapter, we give the reader an in depth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters. We highlight design decisions and practical issues that must be addressed in order to use \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest. As a pedagogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.

@@

The Earth's magnetosphere is a region around the planet where the Earth's own magnetic field dominates the magnetic field of outer space. It is a region which is impinged upon constantly by the solar wind. The ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere. It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.

@@

    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}. A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurrence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure. 
\end{enumerate}

Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forecasts for various geomagnetic time series. The principal data source for forecasting of geomagnetic indices is the \emph{OMNI} data set hosted by NASA on the \emph{Space Physics Data Facility} (SPDF).

@@

Probabilistic forecasting is of particular importance in geophysics applications as the end users of forecasts often require confidence bounds on the said forecasts. It is in this context where \emph{Gaussian Processes} become especially attractive due to their inherent probabilistic formulation and tractability of exact inference.

@@

In the machine learning community, $m(.)$ and $K(.,.)$ are known as the \emph{mean function} and \emph{covariance function} or \emph{kernel function} of the process respectively. Giving a closed form expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed in the following notation.
